Quantitative, Multispecies Monitoring at a Continental Scale

### 2025-03-06 older versions/Appendix_plain_AOS_ORL.tex
older versions folder with old .tex.

We developed a state-space framework for modeling hake DNA concentration in the coastal ocean building and modifying the work of Shelton et al. (2022).  State-space models separate the true biological process from the methods used to observe the process. Let $D_H(x,y,d)$ be the true, but unobserved concentration of hake DNA (log DNA $copies$ $\mu L^{-1}$) present at spatial coordinates $\{x,y\}$ (eastings and northings, respectively, in km) and sample depth $d$ (meters). Hereafter we refer to a combination of spatial location and depth as a ``station". Let $\psi_0$ be the intercept representing the mean hake DNA concentration and $\epsilon(x,y,d)$ be a random effect that allows for the effect of spatial location and water depth. We write these as additive on the log-scale,

\begin{align}
  D_H(x,y,d) &= \psi_0 + \epsilon(x,y,d) \\
  \epsilon(x,y,d) &\sim N(0,\sigma_D)
\end{align}

@@

\begin{align} 
 \nu_{sm} &= \left( L_{sm} - L_{Rm}\right) + N_{PCR} \left(\alpha_s - \alpha_{Ri} \right) 
  \pi_{sm} &= \dfrac{e^{\nu_{sm}}}{\sum_{s=1}^S e^{\nu_{sm}}} \\
  \boldsymbol{R}_{m} & \sim Multinomial\left(\boldsymbol{\pi}_{m},T_{m}\right) 
\end{align}

@@

Ole's questions:
Smoothed model is based on E (a single water bottle), not D.
What was used in the making the smooth?  Posterior mean?

\textbf{OWEN.  COULD YOU HELP HERE?}

The above statistical model provides predictions of species-specific DNA concentrations ($E$) at each depth sampled (d = 0, 50, 150, 300 or 500 m). We used the sdmTMB package \cite{anderson2022} to generate distribution models for the concentration of eDNA , as:
$$E_i(xyz) = \tau + \eta(s) + \varepsilon_{d}(s)$$
where x,y,z index longitude, latitude, and depth, respectively, $\tau$ is scalar intercept, $\eta(s)$ is a spatial smooth shared among depths, and $\varepsilon_d(s)$ is a depth-specific spatial smooth. Both $\eta(s)$ and $\varepsilon_d(s)$ are a zero-centered Gaussian random field estimated using the stochastic partial differntial equation (SPDE) approximation (see \cite{anderson2022}). Each species was fitted individually while we created and selected the best mesh for approximating spatial Gaussian random fields as part of the model.

>>>>

We developed a state-space framework for modeling hake DNA concentration in the coastal ocean building and modifying the work of Shelton et al. (2022).  State-space models separate the true biological process from the methods used to observe the process. Let $D_H(x,y,d)$ be the true, but unobserved concentration of hake DNA (log DNA $copies$ $\mu L^{-1}$) present at spatial coordinates $\{x,y\}$ (eastings and northings, respectively, in km) and sample depth $d$ (meters). Hereafter we refer to a combination of spatial location and depth as a ``station" and index station using $j$ (thus,$D_{Hj} = D_H(x,y,d)$ . Let $\psi_0$ be the intercept representing the mean hake DNA concentration and $\epsilon_j$ be a random effect that allows for the effect of that station. We write these as additive on the log-scale,

\begin{align}
  D_{Hj} &= \psi_0 + \epsilon_j \\
  \epsilon_j &\sim N(0,\sigma_D)
\end{align}

@@

\begin{align} 
 \nu_{sm} &= \left( L_{sm} - L_{Rm}\right) + N_{PCR} \left(\alpha_s - \alpha_{Ri} \right) \\
  \pi_{sm} &= \dfrac{e^{\nu_{sm}}}{\sum_{s=1}^S e^{\nu_{sm}}} \\
  \boldsymbol{R}_{m} & \sim Multinomial\left(\boldsymbol{\pi}_{m},T_{m}\right) 
\end{align}

@@

The above statistical model provides predictions of species-specific DNA concentrations ($E$) at each depth sampled (d = 0, 50, 150, 300 or 500 m). We used the sdmTMB package \cite{anderson2022} to generate distribution models for the concentration of eDNA , as: $$E_i(xyz) = \tau + \eta(xy) + \varepsilon_{d}(xyz)$$ where x,y,z index longitude, latitude, and depth, respectively. We used posterior means of $E_i$ from the joint statistical model above as the dependent variable.  $\tau$ is scalar intercept. Both $\eta(xy)$ and $\varepsilon_d(xyz)$ are species- and location-specific realizations of Gaussian Markov random fields, $\eta \sim MVN(O,\sum_\eta)$ and $\varepsilon_d \sim MVN(O,\sum_{\varepsilon_d})$ (see \cite{anderson2022}).  $\eta$ is a depth-invariant spatial field, while $\varepsilon_d$ is a spatial field specific to each depth category. In simple terms, each station-specific estimate of log eDNA concentration is estimated as a sum of the scalar intercept (that can be zero) and two spatial fields, one that is common across depths and one that is depth-specific. $\eta$ estimates the smooth, two-dimensional (latitude/longitude) spatial pattern in estimated eDNA, while $\varepsilon_d$ modulates that estimate for different sampling depths. The $\varepsilon_d$ fields were allowed to be independent and identically distributed (IID), and therefore there was no forced correlation across depths beyond that provided by the common spatial field $\eta$.

Estimation of large, multi-level spatiotemporal models can strain computational power, in part because of the need in spatial models to invert large covariance matrices between the data. We use the Stochastic Partial Differential Equation (SPDE) model, implemented through the fmesher R package (Lindgren et al. 2024; 10.32614/CRAN.package.fmesher) to approximate the latent spatial process and its variance by using a triangulated mesh (Lindgren et al. 2011). The SPDE approach drastically reduces computation time by using a sparse covariance matrix to approximate Gaussian Markov random fields like $\eta$ and $\varepsilon_d$. Model estimates and the degree of spatial smoothing can be affected by the details of mesh construction, and we therefore produced a range of meshes with different complexities (i.e., varying numbers of triangles, triangle sizes, and therefore vertices) to test for each species.

Each species was fitted individually using the equation above. We tested 18 candidate meshes for each species, increasing in complexity from 33 to 205 vertices, as well as allowing each model to include or exclude the scalar intercept $\tau$. Each candidate model for each species was fit by maximum marginal likelihood, implemented in the sdmTMB package in R ( \cite{anderson2022} ). Model selection was carried out by first ensuring proper convergence, then choosing the model for each species that maximized the summed likelihood.

### 2025-03-06 older versions/DAG_v2.tex
older versions folder with old .tex.

\documentclass[landscape]{article}
\usepackage[paperwidth=252mm, paperheight=180mm, margin=0cm]{geometry}
\usepackage{tikz}
\usepackage{amsmath, amsthm, amssymb, mathtools, thmtools}
\usepackage{xcolor}
\definecolor{myBlue}{RGB}{173, 216, 230}
\definecolor{myGrey}{RGB}{200, 200, 200}

@@

        % Arrows
        \draw [arrow] (qPCR_st) -- node[pos=0.1, below] {} (qPCR_en);
        % \begin{array}{c}
        %     \phi, \\ \beta_0, \beta_1, \\ \gamma_0, \gamma_1
        % \end{array}$} (qPCR_en);
        \draw [arrow] (state1) -- node[midway, left] {} (state2H);
        \draw [arrow] (state2H) -- node[midway, right] {} (qPCR_en);
        \draw [dashed,arrow] (state2s) -- node[midway, right] {}  (spatial);
        %\draw [arrow] (state1) -- node[midway, right] {}  (spatial);
        \draw [arrow] (mock) -- node[pos=0.7, above] {$\alpha_s$} (metabarcoding);
        \draw [arrow] (state2H) -- node[midway, right] {} (metabarcoding);
        \draw [arrow] (state2s) -- node[midway, right] {} (metabarcoding);

% Legend: Parameters
\node (parameters) [rectangle, draw, align=left, below=of metabarcoding, text width=9cm, fill=myGrey] at (3.5,-6){
    \textbf{Parameters and derived quantities} \\
    $\psi$ = mean hake eDNA concentration (log(copies/$\mu$L) \\
    %$\delta0, \delta1$ = offset from mean hake \\
    $\delta$ = biological replicate random effect \\
    %$\omega$ = ethanol wash effect \\
    $V_i$ = volume filtered and aliquote dilution offset \\
   % $\mu$ = mean Ct values of qPCR \\
    $\phi$ =  qPCR amplification parameter\\
    $\sigma_{Y}$ = standard deviation of qPCR Ct values \\
    $\beta_{0p}, \beta_{1p}$ = intercept and slope defining mean qPCR - DNA concentration relationship\\
    $\gamma_0, \gamma_1$ = intercept and slope defining  standard deviation of qPCR - DNA concentration relationship \\
    $\alpha_s$ = amplification efficiency relative to reference species \\
    $\pi$ = predicted sequence proportions at the end of PCR \\
    $\zeta$ = ethanol wash effect \\
    $\tau$ = standard deviation among biological replicates\\
};

@@

% Legend: Data
\node (data) [rectangle, draw, align=left, below=of subscripts, text width=8cm, fill=myGrey] at (12.5,-6){
    \textbf{Observations } \\
    $Z$ = qPCR amplification (yes=1; no=0) \\
    $Y$ = qPCR Ct values \\
    $K$ = known DNA conc. (log(copies/$\mu$L))\\
    $L$ = known DNA conc. (log(copies/$\mu$L))\\
    $W$ = metabarcoding sequencing reads \\
    $T$ = total number sequencing reads \\
    $V$ = volume filtered and aliquote dilution offset \\
    $I$ = aliquote dilution factor \\
};

% Legend: Data
\node (States of Interest) [rectangle, draw, align=left, below=of subscripts, text width=8cm, fill=myGrey] at (12.5,-10.25){
    \textbf{States } \\
    $D$ = DNA conc. at a station-depth (log(copies/$\mu$L)) \\
    $E$ = DNA conc. in a biological replicate (log(copies/$\mu$L)) \\
    $F$ = Smoothed DNA conc. (copies/$L$)\\
    };

>>>>

\documentclass[landscape]{article}
\usepackage[paperwidth=250mm, paperheight=200mm, margin=0cm]{geometry}
\usepackage{tikz}
\usepackage{amsmath, amsthm, amssymb, mathtools, thmtools}
\usepackage{xcolor}
\definecolor{myBlue}{RGB}{173, 216, 230}
\definecolor{myGrey}{RGB}{200, 200, 200}

@@

        % Arrows
        \draw [arrow] (qPCR_st) -- node[pos=0.1, below] {} (qPCR_en);
        % \begin{array}{c}
        %     \phi, \\ \beta_0, \beta_1, \\ \gamma_0, \gamma_1
        % \end{array}$} (qPCR_en);
        \draw [arrow] (state1) -- node[midway, left] {$$} (state2H);
        \draw [arrow] (state2H) -- node[midway, right] {} (qPCR_en);
        \draw [dashed,arrow] (state2s) -- node[midway, right] {}  (spatial);
        %\draw [arrow] (state1) -- node[midway, right] {}  (spatial);
        \draw [arrow] (mock) -- node[pos=0.7, above] {$\alpha_s$} (metabarcoding);
        \draw [arrow] (state2H) -- node[midway, right] {$$} (metabarcoding);
        \draw [arrow] (state2s) -- node[midway, right] {$$} (metabarcoding);

% Legend: Parameters
\node (parameters) [rectangle, draw, align=left, below=of metabarcoding, text width=9cm, fill=myGrey] at (3.5,-6){
    \textbf{Parameters and derived quantities} \\
    $\psi$ = mean hake eDNA concentration \\
    %$\delta0, \delta1$ = offset from mean hake \\
    $\delta$ = biological replicate random effect \\
    %$\omega$ = ethanol wash effect \\
    $V_i$ = volume filtered and aliquote dilution offset \\
   % $\mu$ = mean Ct values of qPCR \\
    $\phi$ =  qPCR amplification parameter\\
    $\sigma_{Y}$ = standard deviation of qPCR Ct values \\
    $\beta_{0p}, \beta_{1p}$ = intercept and slope defining mean qPCR - DNA concentration relationship\\
    $\gamma_0, \gamma_1$ = intercept and slope defining  standard deviation of qPCR - DNA concentration relationship \\
    $\alpha_s$ = amplification efficiency relative to reference species \\
    $\pi$ = predicted sequence proportions at the end of PCR \\
    $\zeta$ = ethanol wash effect \\
    $\tau$ = standard deviation among biological replicates\\
};

@@

% Legend: Data
\node (data) [rectangle, draw, align=left, below=of subscripts, text width=8cm, fill=myGrey] at (12.5,-6){
    \textbf{Observations } \\
    $Z$ = qPCR amplification (yes=1; no=0) \\
    $Y$ = qPCR Ct values \\
    $K$ = known DNA conc. (log(copies/$\mu$L))\\
    $L$ = known DNA conc. (log(copies/$\mu$L))\\
    $W$ = metabarcoding sequencing reads \\
    $T$ = total number sequencing reads \\
};

% Legend: Data
\node (States of Interest) [rectangle, draw, align=left, below=of subscripts, text width=8cm, fill=myGrey] at (12.5,-9.25){
    \textbf{States } \\
    $D$ = DNA conc. at a station-depth (log(copies/$\mu$L)) \\
    $E$ = DNA conc. in a biological replicate (log(copies/$\mu$L)) \\
    $F$ = Smoothed DNA conc. (copies/$L$)\\
    };

### 2025-03-06 older versions/Manuscript_plain_v1.tex
older versions folder with old .tex.

We have shown an example of the substantial value of multi-species eDNA analysis for fisheries and other natural-resource management questions \cite{ledger2024,stoeckle2024}, however, molecular patterns alone are unlikely to drive quantitative natural-resources decisions, in part because eDNA generally does not provide auxiliary information important for managing populations (e.g., information on size- or age structure, fecundity, sex, or disease). Moreover, the eDNA sampling we describe here depends upon survey vessels that can reach and sample from deep water – survey vessels that exist to carry out traditional trawl-based surveys.

Conventional fish surveys, such as trawling or mark-recapture studies, typically rely on a closure assumption, treating a fish population as if it were static and confined during the sampling window \cite{bailey2014}. In reality, fish can move in between sampling location within the study area, which may bias abundance estimates  \cite{rota2009}. By contrast, fish continuously shed DNA into their surroundings, creating a dynamic signal of presence. As the shedded DNA decays, disperses, and degrades over time, fresher DNA remains at higher concentrations while older DNA persists at lower concentrations \cite{thomsen2015,jerde2011}. This cumulative, time-integrated signal allows eDNA surveys to circumvent the closure assumption and provide a more accurate representation of fish presence and activity.

Perhaps most notably, management decisions happen in units of mass or abundance, not in units of eDNA concentration. Relating observed eDNA concentrations to species mass requires data external to the eDNA signal \cite{guri2024a}, and in particular, requires that we understand proportionalities between eDNA concentration and biomass that are likely to vary among species. We therefore see eDNA analysis as strengthening and extending existing surveys, with different data streams complementing one another.

>>>>

We have shown an example of the substantial value of multi-species eDNA analysis for fisheries and other natural-resource management questions \cite{ledger2024,stoeckle2024}, however, molecular patterns alone are unlikely to drive quantitative natural-resources decisions, in part because eDNA generally does not provide auxiliary information important for managing populations (e.g., information on size- or age structure, fecundity, sex, or disease). Moreover, the eDNA sampling we describe here depends upon survey vessels that can reach and sample from deep water – survey vessels that exist to carry out traditional acoustic or trawl-based surveys.

A significant difference between  management decisions rely on units of mass or abundance, not in units of eDNA concentration. Relating observed eDNA concentrations to species mass requires data external to the eDNA signal \cite{guri2024a}, and in particular, requires that we understand proportionalities between eDNA concentration and biomass that may vary among species. We therefore see eDNA analysis as strengthening and extending existing surveys, with different data streams complementing one another.

%Conventional fish surveys, such as trawling or mark-recapture studies, typically rely on a closure assumption, treating a fish population as if it were static and confined during the sampling window \cite{bailey2014}. In reality, fish can move in between sampling location within the study area, which may bias abundance estimates  \cite{rota2009}. By contrast, fish continuously shed DNA into their surroundings, creating a dynamic signal of presence. As the shedded DNA decays, disperses, and degrades over time, fresher DNA remains at higher concentrations while older DNA persists at lower concentrations \cite{thomsen2015,jerde2011}. This cumulative, time-integrated signal allows eDNA surveys to circumvent the closure assumption and provide a more accurate representation of fish presence and activity.
